Parallel Hierarchical Clustering using \\ Rank-Two Nonnegative Matrix Factorization \thanks{Identify applicable funding agency here. If none, delete this.}

### 2020-07-16 latex/paper.tex
stashing the change in paper.tex

\begin{abstract}
Nonnegative Matrix Factorization (NMF) is an effective tool for clustering nonnegative data, either for computing a flat partitioning of a dataset or for determining a hierarchy of similarity.
In this paper, we propose a parallel algorithm for hierarchical clustering that uses a divide-and-conquer approach based on rank-two NMF to split a data set into two cohesive parts.
Not only does this approach uncover more structure in the data than a flat NMF clustering, but also rank-two NMF can be computed more quickly than for general ranks, providing comparable overall time to solution.
Our data distribution and parallelization strategies are designed to maintain computational load balance throughout the data-dependent hierarchy of computation while limiting interprocess communication, allowing the algorithm to scale to large dense and sparse data sets.
We demonstrate the scalability of our parallel algorithm in terms of data size (up to 800 GB) and number of processors (up to 80 nodes of the Summit supercomputer), applying the hierarchical clustering approach to hyperspectral imaging and image classification data. 
Our algorithm for Rank-2 NMF scales perfectly on up to 1000s of cores and the entire hierarchical clustering method achieves 5.9x speedup scaling from 10 to 80 nodes on an 800 GB dataset.
\end{abstract}

@@

While standard NMF provides a flat clustering, NMF can also be used as a tool for hierarchical clustering.
Kuang and Park \cite{KP13} propose a method that uses rank-2 NMF to recursively bipartition the samples.
An NMF with $k=2$ determines a row of $\M{H}$ of length 2 for each sample that determines its part membership.
By recursively bipartitioning the samples in each part, the method determines a binary tree such that each sample belongs to a unique leaf and internal nodes of the tree determine higher-level clusters of the samples.
A single latent feature vector for each node can also be used for interpreting clusters.
Du et al. \cite{DKDP17} show that the latent features vectors of the leaves can also be used to initialize a $\M{W}$ within an algorithm for flat NMF, yielding fast convergence and overall speedup compared to initializing the algorithm with random data.
We discuss the hierarchical method in more detail in \Cref{sec:prelim} and \Cref{sec:seqalgs}.

>>>>

\begin{abstract}
Nonnegative Matrix Factorization (NMF) is an effective tool for clustering nonnegative data, either for computing a flat partitioning of a dataset or for determining a hierarchy of similarity.
In this paper, we propose a parallel algorithm for hierarchical clustering that uses a divide-and-conquer approach based on rank-two NMF to split a data set into two cohesive parts.
Not only does this approach uncover more structure in the data than a flat NMF clustering, but also rank-two NMF can be computed more quickly than for general ranks, providing comparable overall time to solution.
Furthermore, the rank-two NMF algorithm can be parallelized efficiently using a computational kernel similar to matrix-vector multiplication.
Our data distribution and parallelization strategies are designed to maintain computational load balance throughout the data-dependent hierarchy of computation while limiting interprocess communication, allowing the algorithm to scale to large dense and sparse data sets.
We demonstrate the scalability of our parallel algorithm in terms of data size and number of processors, and apply the hierarchical clustering approach to large (dense) hyperspectral imaging and (sparse) text document data. \GB{On 40 nodes of Summit supercomputer, our algorithm achieves (a) linear speedup for solving the Rank-2 Non-negative Least squares kernels and (b) achieves 14x speedup on synthetic matrices. We also demonstrate that in the case of real world matrices, the algorithm scales to 80 summit nodes.} 
\end{abstract}

@@

Hierarchical clustering is the process of recursively breaking a group into two similar groups until every sample is a cluster by itself. That is., we consider all the data points as a single cluster and in each iteration, the data points in the current cluster are split into two separate clusters till there is only one sample in a cluster.
While standard NMF is interpreted as a flat clustering, it can also be extended for hierarchical clustering.
Kuang and Park \cite{KP13} propose a method that uses rank-2 NMF to recursively bipartition the samples.
An NMF with $k=2$ determines a row of $\M{H}$ of length 2 for each sample that determines its part membership.
By recursively bipartitioning the samples in each part, the method determines a binary tree such that each sample belongs to a unique leaf and internal nodes of the tree determine higher-level clusters of the samples.
A single latent feature vector for each node can also be used for interpreting clusters.
Similarly, Du et al. \cite{DKDP17} show that the latent features vectors of the leaves can also be used to initialize a $\M{W}$ within an algorithm for flat NMF, yielding fast convergence and overall speedup compared to initializing the algorithm with random data.
We discuss the hierarchical method in more detail in \Cref{sec:prelim} and \Cref{sec:seqalgs}.
